Ricci-Ollivier Curvature of the Rooted Phylogenetic Subtree-Prune-Regraft Graph

### 2015-07-01 curvature.tex
graph construction revisions

\section{Efficient algorithms for computing and sampling the rSPR graph}

\subsection{Computing the rSPR graph on $n$ leaves in $\OhOf{mn^3}$-time}
\label{sec:computing_treespace}

To study the full rSPR graph for a fixed number of leaves it is necessary to have an efficient means of constructing it.
The previous best algorithm for this problem requires $\OhOf{m^2 n}$ time, where $m$ is the number of trees in the graph and $n$ the number of leaves~\cite{Whidden2015-yi}.
Here we reduce that time to $\OhOf{mn^3}$.
Note that for the full rSPR graph, $m$ is the rapidly growing function $(2n-3)!!$, that is, $3 \cdot 5 \cdot \ldots \cdot (2n-3)$, and this is therefore a significant improvement.

@@

To quickly compute dense rSPR graphs (those containing a significant portion of the full rSPR graph) we avoid pairwise computations by successively adding trees to the graph and \cuttable{efficiently }identifying their neighbors: \todo{make this more concrete and CSy, push the proofs to the supplemental material.}
\begin{enumerate}[label={\arabic*}.]
	\item Let $G$ be an empty graph.
	\item For each of the $m$ trees in a specified but arbitrary order:
		\begin{enumerate}
			\item Add a vertex $i$ to the graph representing the current tree $T$ where $i$ is $T$'s order index.
			\item For each of the $\OhOf{n^2}$ neighbors of $T$:
				\begin{enumerate}
					\item If the current neighbor $S$'s index is in $G$ then add an edge $(j,i)$ to the graph, where $j$ is $S$'s order index.
				\end{enumerate}
		\end{enumerate}
\end{enumerate}

@@

	We implement the graph with an adjacency list representation with integer-labelled vertices that supports $\OhOf{\log n}$ edge insertions and lookups (with e.g. red-black trees~\cite{guibas1978dichromatic}, as the vertex degrees are $\OhOf{n^2}$).
	As described above, the integer labels are simply the order of the input trees.
	Adding the vertices to the graph requires $\OhOf{m}$-time, as they are added in ascending order to the end of the vertex list, which can be stored as a fixed-size array.
	Adding the $\OhOf{mn^2}$ edges to the graph requires $\OhOf{mn^2 \log n}$-time.
	Enumerating the neighbors of $T$ requires $\OhOf{n^3}$-time for each $T$, for a total of $\OhOf{mn^3}$-time.
	We discuss below, in Section~\ref{sec:random_walks} how to do so efficiently without considering duplicate neighbors.
	We store the tree to index mappings for current vertices of $G$ in a trie~\cite{fredkin1960trie} using the $\OhOf{n}$-length Newick \cite{wiki:newick} representation of the tree as a string.
	These representations are made unique by ordering the tree so that leftmost subtrees contain the smallest alphanumeric label of descendants.
	This requires only $\OhOf{n}$-time for each tree (i.e. a total of $\OhOf{mn^3}$-time) using a standard nodes-and-pointers representation of the tree and assuming integer leaf labels (a simple $\OhOf{mn\log n}$ leaf preprocessing step could be applied to extend this procedure to phylogenetic trees with string labels).
	Similarly, it takes $\OhOf{n}$-time to determine the index of each of the $\OhOf{mn^2}$ $S$'s.
	Therefore the graph can be constructed in $\OhOf{mn^3}$-time.
\end{proof}

This procedure has been implemented in the C++ program \texttt{dense\_spr\_graph} of the software package \texttt{spr\_neighbors}~\cite{spr_neighbors}, which outputs an edge list format graph suitable for input to other software.
The construction procedure reduced the time required to compute the 10,395-vertex 7-taxon rSPR graph from 2,104.68 seconds to 12.71 seconds on an Intel Core 2 Duo E7500 desktop running Ubuntu 14.04.
Moreover, although we do not study the 135,135-vertex 8-taxon rSPR graph in this paper, our algorithm required only 303.45 seconds to construct it on the same hardware.
Constructing the 8-taxon rSPR graph using the previous method required 377,395 seconds (more than 4 days), and thus that method is infeasible for constructing larger rSPR tree graphs.
Thus, we believe our fast graph construction procedure will itself be useful for further studies of rSPR graph subsets similar to~\cite{Whidden2015-yi}.

@@

\begin{corollary}
	The rSPR neighbors of a tree $T$ can be enumerated in $\OhOf{n^3}$-time.
\end{corollary}

>>>>

\section{Efficient algorithms for computing and sampling rSPR graphs}

\subsection{Computing the rSPR graph of $m$ trees with $n$ leaves in $\OhOf{mn^3}$-time}
\label{sec:computing_treespace}

To study the full rSPR graph for a fixed number of leaves it is necessary to have an efficient means of constructing it.
The previous best algorithm for this problem requires $\OhOf{m^2 n}$ time, where $m$ is the number of trees in the graph and $n$ the number of leaves~\cite{Whidden2015-yi}.
Here we reduce that time to $\OhOf{mn^3}$.
Note that for the full rSPR graph, $m$ is the rapidly growing function $(2n-3)!!$, that is, $3 \cdot 5 \cdot \ldots \cdot (2n-3)$, and this is therefore a significant improvement in practice, as we demonstrate below.

@@

The key to our efficient algorithm for quickly computing dense rSPR graphs (those containing a significant portion of the full rSPR graph) lies in avoiding the pairwise comparison of non-adjacent trees and thereby shaving off an $\OhOf{m}$ factor.
The input to our algorithm is a set $\mathcal{T}$ of phylogenetic trees in the $\OhOf{n}$-length Newick \cite{wiki:newick} representation of each tree as a string.
These representations are made unique by ordering each tree so that leftmost subtrees contain the smallest alphanumeric label of descendants.
We construct a mapping from each tree $T_i$ to its order index in this list $i$.
We then begin with an empty graph $G$.
For each tree $T_i$, we first add a vertex $i$ to the graph and then use Corollary~\ref{cor:enumerate_neighbors} to enumerate the $\OhOf{n^2}$ neighbors of $T_i$ in the rSPR graph in $\OhOf{n^3}$-time. 
This efficient enumeration procedure is the key step required to achieve our desired running time of $\OhOf{mn^3}$.
We use the tree to index mappings to determine whether these trees already reside in the tree and, if so, add an edge in the graph from $T_i$ to each such neighbor $T_j$.
The high-level steps are as follows, and we show in Theorem~\ref{thm:construct_graph} that this algorithm is correct and can be implemented to run in the stated time.

\textsc{Construct-rSPR-Graph($\mathcal{T}$)}
\begin{enumerate}[label={\arabic*}.]
	\item Let $G$ be an empty graph.
	\item Let $M$ be a mapping from trees to integers.
	\item Let $i = 0$.
	\item For each of the $m$ trees:
		\begin{enumerate}
			\item Add a vertex $i$ to $G$ representing the current tree $T_i$.
			\item Add $T_i \rightarrow i$ to $M$.
			\item For each of the $\OhOf{n^2}$ neighbors of $T_i$, enumerated using \textsc{Enumerate-rSPR-Neighbors($T_i$)}:
				\begin{enumerate}
					\item If the current neighbor $T_j$ is in $M$ then add an edge $(i,M[T_j])$ to $G$.
				\end{enumerate}
		\item $i = i + 1$.
		\end{enumerate}
\end{enumerate}

@@

	We implement the graph with an adjacency list representation with integer-labelled vertices that supports $\OhOf{\log n}$ edge insertions and lookups (with e.g. red-black trees~\cite{guibas1978dichromatic}, as the vertex degrees are $\OhOf{n^2}$).
	As described above, the integer labels are simply the order of the input trees.
	Adding the vertices to the graph requires $\OhOf{m}$-time, as they are added in ascending order to the end of the vertex list, which can be stored as a fixed-size array.
	Adding the $\OhOf{mn^2}$ edges to the graph requires $\OhOf{mn^2 \log n}$-time.
	Enumerating the neighbors of $T_i$ requires $\OhOf{n^3}$-time for each $T_i$, for a total of $\OhOf{mn^3}$-time.
	We discuss below, in Section~\ref{sec:random_walks} how to do so efficiently without considering duplicate neighbors.
	We store the tree to index mappings for current vertices of $G$ in a trie~\cite{fredkin1960trie} using Newick representation.
	This requires only $\OhOf{n}$-time for each tree (i.e. a total of $\OhOf{mn^3}$-time) using a standard nodes-and-pointers representation of the tree and assuming integer leaf labels (a simple $\OhOf{mn\log n}$ leaf preprocessing step could be applied to extend this procedure to phylogenetic trees with string labels).
	Similarly, it takes $\OhOf{n}$-time to determine the index of each of the $\OhOf{mn^2}$ considered neighbors.
	Therefore the graph can be constructed in $\OhOf{mn^3}$-time, as claimed.
\end{proof}

We implemented this procedure in the C++ program \texttt{dense\_spr\_graph} of the software package \texttt{spr\_neighbors}~\cite{spr_neighbors}, which outputs an edge list format graph suitable for input to other software.
The construction procedure reduced the time required to compute the 10,395-vertex 7-taxon rSPR graph from 2,104.68 seconds to 12.71 seconds on an Intel Core 2 Duo E7500 desktop running Ubuntu 14.04.
Moreover, although we do not study the 135,135-vertex 8-taxon rSPR graph in this paper, our algorithm required only 303.45 seconds to construct it on the same hardware.
Constructing the 8-taxon rSPR graph using the previous method required 377,395 seconds (more than 4 days), and thus that method is infeasible for constructing larger rSPR tree graphs.
Thus, we believe our fast graph construction procedure will itself be useful for further studies of rSPR graph subsets similar to~\cite{Whidden2015-yi}, as the algorithm can construct rSPR graphs for any given subset of trees.

@@

\begin{corollary}
\label{cor:enumerate_neighbors}
	The rSPR neighbors of a tree $T$ can be enumerated in $\OhOf{n^3}$-time.
\end{corollary}
